---
title: "Distance sampling targets pipeline"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# set up
library(knitr)
opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)
opts_knit$set(root.dir = here::here())
```

```{r}
# packages
library(tidyverse)
library(targets)
library(Distance)

library(INBOtheme)
theme_set(theme_inbo(transparent = TRUE))

# Conflicts
conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)

# Paths
mbag_dir <- rprojroot::find_root_file(criterion = rprojroot::is_git_root)
targets_store <- file.path(mbag_dir, "source", "targets", "distance_sampling",
                           "_targets")

# Source functions
source(file.path(mbag_dir, "source", "R", "tar_read_ds.R"))
source(file.path(mbag_dir, "source", "R", "plot_pdf_per_cat.R"))
source(file.path(mbag_dir, "source", "R", "plot_detection_curve.R"))
```

# Doel

# Densiteitsschattingen
## Methoden

## Veldleeuwerik
### Data exploratie

### Model selectie

```{r}
final_model <- tar_read_ds("model_selection", "Veldleeuwerik", 2024)
f <- final_model$ddf$ds$aux$ddfobj$scale$formula
key <- ifelse(
  final_model$ddf$ds$aux$ddfobj$type == "hn",
  "half-normal",
  "hazard-rate"
)
```

We select the model with `r key` key function and scale parameter formula `r f`.

```{r}
tar_read_ds("aic_comparison", "Veldleeuwerik", 2024) %>%
  select(-species, -year) %>%
  filter(`Delta AIC` <= 10) %>% # only show low AIC models
  kable()
```

### Model fit

We look at the fit of the best model.

```{r}
par(mfrow = c(1, 2))
gof_ds(final_model)
plot(final_model, pdf = TRUE, showpoints = TRUE)
par(mfrow = c(1, 1))
```

```r
plot_pdf_per_cat(final_model)
```

The detection curve(s) look like this:

```{r}
plot_detection_curve(
  final_model,
  show_data = TRUE,
  plot_average_fit = TRUE,
  plot_covariate_fit = TRUE
) +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 100, by = 0.25)) +
  labs(x = "Distance (m)", y = "Detection probability") +
  theme_minimal() +
  theme(
    legend.background = element_rect(fill = "white", color = "darkgrey"),
    legend.margin = margin(6, 6, 6, 6),
    legend.position = "bottom"
  )
```

### Resultaten

```{r}
y <- 2024
```

Detection probabilities.

```{r}
tar_read_ds("detection_probabilities", "Veldleeuwerik") %>%
  filter(year == y) %>%
  select(-species, -year) %>%
  drop_na() %>%
  relocate("estimate_p", "se_p", "ll_beta", "ul_beta", .after = last_col()) %>%
  kable()
```

#### Stratum

Abundances.

```{r}
abundances_stratum_target <- tar_read_ds("abundances_stratum", "Veldleeuwerik")
abundances_stratum <- abundances_stratum_target %>%
  filter(
    if (y == 2023) {
      year == y & grepl("^Zandleem|^Leem", Label)
    } else {
      year == y
    }
  ) %>%
  select(-species, -year) %>%
  rename("stratum" = "Label", "estimate" = "Estimate")
abundances_stratum %>%
  kable()
```

Densities.

```{r}
densities_stratum_target <- tar_read_ds("densities_stratum", "Veldleeuwerik")
densities_stratum <- densities_stratum_target %>%
  filter(
    if (y == 2023) {
      year == y & grepl("^Zandleem|^Leem", Label)
    } else {
      year == y
    }
  ) %>%
  select(-species, -year) %>%
  rename("stratum" = "Label", "estimate" = "Estimate")
densities_stratum %>%
  kable()
```

```{r}
densities_stratum %>%
  separate(stratum, into = c("regio", "openheid", "sbp"), sep = " - ") %>%
  mutate(stratum = paste(openheid, sbp, sep = " - ")) %>%
  ggplot(aes(x = openheid, y = estimate, colour = sbp)) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lcl, ymax = ucl),
    width = 0.2,
    position = position_dodge(width = 0.5)
  ) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(y = "breeding pairs / 100 ha", x = "") +
  facet_wrap(~regio, ncol = 2, scales = "free")
```

Over the years:

```{r}
densities_stratum_target %>%
  filter(
    year == 2023 & grepl("^Zandleem|^Leem", Label) |
    year != 2023
  ) %>%
  rename("stratum" = "Label", "estimate" = "Estimate") %>%
  filter(stratum != "Total") %>%
  separate(stratum, into = c("regio", "openheid", "sbp"), sep = " - ") %>%
  mutate(stratum = paste(openheid, sbp, sep = " - ")) %>%
  ggplot(aes(x = year, y = estimate, colour = stratum)) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lcl, ymax = ucl),
    width = 0.2,
    position = position_dodge(width = 0.5)
  ) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(y = "breeding pairs / 100 ha", x = "") +
  facet_wrap(~regio, ncol = 2, scales = "free")
```

#### Regio

Abundances.

```{r}
abundances_region_target <- tar_read_ds("abundances_region", "Veldleeuwerik")
abundances_region <- abundances_region_target %>%
  filter(
    if (y == 2023) {
      year == y & grepl("^Zandleem|^Leem", Label)
    } else {
      year == y
    }
  ) %>%
  select(-species, -year) %>%
  rename("region" = "Label", "estimate" = "Estimate")
abundances_region %>%
  kable()
```

Densities.

```{r}
densities_region_target <- tar_read_ds("densities_region", "Veldleeuwerik")
densities_region <- densities_region_target %>%
  filter(
    if (y == 2023) {
      year == y & grepl("^Zandleem|^Leem", Label)
    } else {
      year == y
    }
  ) %>%
  select(-species, -year) %>%
  rename("region" = "Label", "estimate" = "Estimate")
densities_region %>%
  kable()
```

```{r}
densities_region %>%
  filter(region != "Total") %>%
  ggplot(aes(x = region, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lcl, ymax = ucl), width = 0.2, linewidth = 1) +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 100, 2)) +
  labs(y = "breeding pairs / 100 ha", x = "")
```

Over the years:

```{r}
densities_region_target %>%
  filter(
    year == 2023 & grepl("^Zandleem|^Leem", Label) |
    year != 2023
  ) %>%
  rename("region" = "Label", "estimate" = "Estimate") %>%
  filter(region != "Total") %>%
  ggplot(aes(x = year, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = lcl, ymax = ucl),
    width = 0.2,
  ) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(y = "breeding pairs / 100 ha", x = "") +
  facet_wrap(~region, ncol = 2, scales = "free")
```

# Spatiale predicties
## Methoden

## Veldleeuwerik
### Selecteer responsdistributie

### Selecteer spatiale spline
#### Isotropie


#### Basisdimentie

### Model fit

### Predicties
