# Remote sensing–indicatoren op basis van optische satelliettijdsreeksen {#sec-remote-sensing}
## Inleiding

- zie aparte google doc

## Methodiek

- zie aparte google doc

## Resultaten

```{r}
#| include: FALSE
# set up
library(knitr)
opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)
```

```{r}
# packages
library(tidyverse)
library(targets)
library(terra)
library(sf)

library(INBOtheme)
theme_set(theme_inbo(transparent = TRUE))

# Conflicts
conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)

# Paths
mbag_dir <- rprojroot::find_root_file(criterion = rprojroot::is_git_root)
store <- file.path(
  mbag_dir, "source", "targets", "data_preparation", "_targets"
)
data_path <- file.path(mbag_dir, "data", "remote_sensing")
```

### Rasterlagen

We krijgen per telpunt rasterlagen met NDVI- en BSI-waarden om de tien dagen van 1 februari tot en met 1 september (vanaf 2022). Dit is voor een beperkte periode gevisualiseerd voor telpunt Ol_12833.8 (Oostelijke leemstreek, HOL, binnen SBP) in @fig-raster-indices. We zien duidelijke verschillen in indexwaarden tussen verschillende percelen. We zien ook verschillende trends in naakte bodem per perceel doorheen de tijd (bij sommige percelen stijgt de index, terwijl deze bij andere daalt). We zien gelijkaardige trends voor NDVI als voor BSI. De figuren geven aan dat dergelijke gegevens nuttig zijn, omdat ze gedetailleerde spatiale en temporele informatie opleveren. Hierdoor kunnen veranderingen in vegetatie en bodemtoestand op een fijnmazig niveau worden opgevolgd, wat waardevolle input biedt voor verdere analyses.

```{r}
#| output: false
plots <- tibble(
  plotnaam = c("Ol_12833.8", "Ol_7536.6", "Ol_9077.8", "Ol_9797.2"),
  regio = "Oostelijke leemstreek",
  openheid = rep(c("HOL", "OL"), each = 2),
  sbp = rep(c("binnen", "buiten"), 2)
)

# Get NDVI rasters
rast_ndvi_list <- apply(plots, 1, function(row) {
  # Create file path
  plotnaam <- row[["plotnaam"]]
  file_name <- paste("multiband_raster_ndvi", plotnaam, "sg.tif", sep = "_")

  # Read tif file
  raster <- rast(file.path(data_path, file_name))

  # Select date range
  raster[[grepl("^2024-0[^12]", names(raster))]]
})
names(rast_ndvi_list) <- plots$plotnaam

# Get BSI rasters
rast_bsi_list <- apply(plots, 1, function(row) {
  # Create file path
  plotnaam <- row[["plotnaam"]]
  file_name <- paste("multiband_raster_bsi", plotnaam, "sg.tif", sep = "_")

  # Read tif file
  raster <- rast(file.path(data_path, file_name))

  # Select date range
  raster[[grepl("^2024-0[^12]", names(raster))]]
})
names(rast_bsi_list) <- plots$plotnaam

#-----------------------------
# NDVI PLOT
#-----------------------------
# Compute global range
ranges_ndvi <- sapply(
  rast_ndvi_list, function(r) range(values(r), na.rm = TRUE)
)
rng <- c(min(ranges_ndvi[1, ]), max(ranges_ndvi[2, ]))

# Define colors and breaks
pal <- rev(map.pal("elevation", 50))
brks <- pretty(rng, n = 5)

# Save original par settings to restore later
op <- par(no.readonly = TRUE)

# Save NDVI plot to a temporary file
ndvi_file <- tempfile(fileext = ".png")
png(ndvi_file, width = 1800, height = 1400, res = 200)

# Plot the raster (without legend)
plot(rast_ndvi_list[[1]], col = pal, range = rng, legend = FALSE, maxnl = 3 * 5)

# Add legend in bottom-right corner
par(fig = c(0.10, 0.95, 0, 0.2), new = TRUE, mar = c(0, 0, 0, 0))
fields::image.plot(
  legend.only = TRUE,
  zlim = rng,
  col = pal,
  legend.args = list(text = "NDVI", side = 3, line = 1),
  axis.args = list(at = brks, labels = format(brks, digits = 2)),
  legend.shrink = 0.7
)
dev.off()

# Restore original graphical parameters safely
par(op)

#-----------------------------
# BSI PLOT
#-----------------------------
# Compute global range
ranges_bsi <- sapply(rast_bsi_list, function(r) range(values(r), na.rm = TRUE))
rng <- c(min(ranges_bsi[1, ]), max(ranges_bsi[2, ]))

# Define colors and breaks
pal <- map.pal("elevation", 50)
brks <- pretty(rng, n = 5)

# Save original par settings to restore later
op <- par(no.readonly = TRUE)

bsi_file <- tempfile(fileext = ".png")
png(bsi_file, width = 1800, height = 1400, res = 200)

# Plot the raster (without legend)
plot(rast_bsi_list[[1]], col = pal, range = rng, legend = FALSE, maxnl = 3 * 5)

# Add legend in bottom-right corner
par(fig = c(0.10, 0.95, 0, 0.2), new = TRUE, mar = c(0, 0, 0, 0))
fields::image.plot(
  legend.only = TRUE,
  zlim = rng,
  col = pal,
  legend.args = list(text = "BSI", side = 3, line = 1),
  axis.args = list(at = brks, labels = format(brks, digits = 2)),
  legend.shrink = 0.7
)
dev.off()

# Restore original graphical parameters safely
par(op)
```

```{r}
#| label: fig-raster-indices
#| fig-cap: "Rasterbeelden van landbouwpercelen rond telpunt Ol_12833.8 van
#|  02-03-2024 tot en met 20-07-2024.
#|  A. NDVI: een lagere waarde betekend minder bedekking.
#|  B. BSI: een hogere waarde betekend minder bedekking."
#| fig-height: 12.5
#| fig-width: 8
#-----------------------------
# Combine with cowplot
#-----------------------------
p_ndvi <- cowplot::ggdraw() + cowplot::draw_image(ndvi_file)
p_bsi  <- cowplot::ggdraw() + cowplot::draw_image(bsi_file)

# Combine side by side with labels
combined <- cowplot::plot_grid(
  p_ndvi, p_bsi,
  ncol = 1,
  labels = c("A.", "B."),
  label_size = 20,
  label_fontface = "bold",
  scale = c(0.9, 0.9)
)

# Show combined figure
combined
```

### Tijdsreeksen

```{r}
# Read time series data from CSV
vec <- c(ndvi = "ndvi", bsi = "bsi")
time_series_csv <- lapply(vec, function(indicator) {
  file_name <- paste0("time_series_mas_", indicator, ".csv")
  path <- file.path(data_path, file_name)
  
  read_csv(path, show_col_types = FALSE)[, -1]
})

# Filter and prepare data
time_series_data <- lapply(time_series_csv, function(df) {
  # Rules for different indicator types
  if ("ndvi" %in% names(df)) {
    y_var <- "ndvi"
    cutoff <- 0.3
    rule <- `<`
  } else {
    y_var <- "bsi"
    cutoff <- 0.021
    rule <- `>`
  }

  names(df) <- tolower(names(df)) # Lower column names

  # Output dataframe
  df %>%
    # Create date variables
    mutate(
      day = day(time),
      month = month(time),
      monthday = format(time, "%m%d"),
      year = year(time)
    ) %>%
    # Filter data
    drop_na() %>%
    filter(
      monthday >= "0201",
      monthday <= "0901"
    ) %>%
    # Describe trends
    mutate(
      naakte_bodem = ifelse(exec(rule, !!sym(y_var), cutoff),
                            "onbedekt", "bedekt"),
    ) %>%
    arrange(pointid, year, time) %>%
    group_by(ref_id, year) %>%
    mutate(
      trend_raw = paste0(rle(naakte_bodem)$values, collapse = " → "),
      trend = factor(
        case_when(
          trend_raw == "onbedekt → bedekt" ~ "toename bedekking",
          trend_raw == "bedekt → onbedekt" ~ "afname bedekking",
          trend_raw == "onbedekt → bedekt → onbedekt" ~ "parabool",
          trend_raw == "bedekt → onbedekt → bedekt" ~ "omgekeerde parabool",
          trend_raw == "bedekt" ~ "altijd bedekt",
          trend_raw == "onbedekt" ~ "altijd onbedekt",
          TRUE ~ "complex patroon"
        ),
        levels = c("altijd bedekt", "altijd onbedekt", "toename bedekking",
                   "afname bedekking", "parabool", "omgekeerde parabool",
                   "complex patroon")
      )
    ) %>%
    ungroup()
})
```

```{r}
# Plot indicator time series for a plot
plot_time_series <- function(
  df,
  loc_name,
  colour_var,
  shape_var = NULL,
  legend_colour = "Legende",
  legend_shape = "Legende"
) {
  # Rules for different indicator types
  if ("ndvi" %in% names(df)) {
    y_axis_title <- "NDVI"
    y_var <- "ndvi"
    cutoff <- 0.3
  } else {
    y_axis_title <- "BSI"
    y_var <- "bsi"
    cutoff <- 0.021
  }
  
  # Prepare data
  data <- df %>%
    # Get plot
    filter(pointid %in% loc_name) %>%
    # Create variable to set x-axis limits
    mutate(
      monthday_date = as.Date(
        paste0("2000-", substr(monthday, 1, 2), "-", substr(monthday, 3, 4))
      )
    )
  
  # Return simple plot
  if (is.null(shape_var)) {
    # Create output
    p <- data %>%
      ggplot(aes(x = monthday_date, y = !!sym(y_var), group = ref_id)) +
      # Create time series
      geom_line(alpha = 0.5, linetype = 5) +
      geom_point(aes(colour = !!sym(colour_var)), size = 2, show.legend = TRUE) +
      # Add reference lines
      geom_hline(yintercept = cutoff, linetype = "dotdash", colour = "black") +
      # Settings
      labs(x = "", y = y_axis_title, colour = legend_colour) +
      scale_x_date(date_breaks = "1 month", date_labels = "%b",
                   limits = as.Date(c("2000-02-01", "2000-09-01"))) +
      scale_colour_discrete(drop = FALSE) +
      facet_wrap(~year, ncol = 1, scales = "fixed")
    return(p)
  }
  
  # Return complex plot
  data %>%
    ggplot(aes(x = monthday_date, y = !!sym(y_var), group = ref_id)) +
    # Create time series
    geom_line(alpha = 0.5, linetype = 5) +
    geom_point(aes(colour = !!sym(colour_var), shape = !!sym(shape_var)),
               size = 2, show.legend = TRUE) +
    # Add reference lines
    geom_hline(yintercept = cutoff, linetype = "dotdash", colour = "black") +
    # Settings
    labs(x = "", y = y_axis_title, colour = legend_colour, shape = legend_shape) +
    scale_x_date(date_breaks = "1 month", date_labels = "%b",
                 limits = as.Date(c("2000-02-01", "2000-09-01"))) +
    scale_shape_manual(
      values = c("bedekt" = 16, "onbedekt" = 4),
      drop = FALSE
    ) +
    scale_colour_discrete(drop = FALSE) +
    facet_wrap(~year, ncol = 1, scales = "fixed")
}
```

```{r}
#| label: fig-tijdsreeks-trend
#| fig-cap: "Tijdsreeks van landbouwpercelen rond telpunt Ol_12833.8 van
#|  1 februari tot en met 1 september.
#|  A. NDVI: afsnijding naakte bodem onder een waarde van 0.3.
#|  B. BSI: afsnijding naakte bodem boven een waarde van 0.021."
#| fig-height: 13
#| fig-width: 8
# Create plots
plot_trend_full <- lapply(
  time_series_data,
  plot_time_series,
  loc_name = plots$plotnaam[1],
  colour_var = "trend",
  legend_colour = "Trend in bedekking",
  shape_var = "naakte_bodem",
  legend_shape = "Naakte bodem"
)

# 1. Extract the shared legend (keep it vertical)
legend_shared <- cowplot::get_legend(
  plot_trend_full$ndvi +
    theme(
      legend.position = "right",
      legend.direction = "vertical"
    )
)

# 2. Remove legends from the individual plots
p_ndvi <- plot_trend_full$ndvi + theme(legend.position = "none")
p_bsi  <- plot_trend_full$bsi  + theme(legend.position = "none")

# 3. Combine the two plots vertically
plots_combined <- cowplot::plot_grid(
  p_ndvi, p_bsi,
  ncol = 1,
  labels = c("A.", "B."),
  label_size = 20,
  label_fontface = "bold",
  scale = c(0.95, 0.95)
)

# 4. Add the shared legend to the right of both
final_plot <- cowplot::plot_grid(
  plots_combined, legend_shared,
  ncol = 2,
  rel_widths = c(1, 0.3)  # adjust legend width
)

# 5. Display final figure
final_plot
```

```{r}
# Read sample information
sample <- tar_read("sample", store = store)
sample_df <- sample %>%
  select(pointid, regio, openheid_klasse, sbp)
```

```{r}
# Read crop type data
years <- 2022:2024

# Helper to read parcel data for one year
read_parcels_year <- function(year) {
  gpkg_path <- file.path(
    mbag_dir, "data", "landbouwgebruikspercelen",
    sprintf("Landbouwgebruikspercelen_%d", year),
    sprintf("Landbouwgebruikspercelen_%d.gpkg", year)
  )
  
  st_read(gpkg_path, quiet = TRUE) %>%
    st_drop_geometry() %>%
    rename_with(tolower) %>%
    mutate(year = year,
           ref_id = as.numeric(ref_id))
}

# Read and combine all years
parcels_all <- map_dfr(years, read_parcels_year)

# Join crop group info
parcels_all_joined <- parcels_all %>%
  inner_join(distinct(time_series_data$ndvi, ref_id, year),
             by = join_by(ref_id, year)) %>%
  select(ref_id, year, gwsnam_h, gwsgrp_h = gwsgrph_lb)
```

```{r}
# Add variables to time series data
time_series_data_full <- lapply(time_series_data, function(df) {
  df %>%
    left_join(sample_df, by = join_by(pointid)) %>%
    left_join(parcels_all_joined, by = join_by(ref_id, year))
})

# Get trend data
trend_data <- lapply(time_series_data_full, function(df) {
  # Rules for different indicator types
  if ("ndvi" %in% names(df)) {
    indicator <- "ndvi"
  } else {
    indicator <- "bsi"
  }
  
  df %>%
    distinct(pointid, ref_id, year, trend_raw, trend, regio,
             openheid_klasse, sbp, gwsnam_h, gwsgrp_h) %>%
    mutate(
      gwsgrp_h = fct_reorder(gwsgrp_h, as.numeric(trend), mean),
      indicator = indicator,
    )
})
```

```{r}
#| label: fig-trend-pairwise
#| fig-cap: "..."
#| out-width: "80%"
pairwise_trends <- bind_rows(trend_data) %>%
  pivot_wider(id_cols = c("pointid", "ref_id", "year"),
              names_from = "indicator", values_from = "trend") %>%
  mutate(
    across(
      c("ndvi", "bsi"),
      ~ case_when(
        grepl("altijd", .x) ~ 0,
        grepl("bedekking", .x) ~ 1,
        grepl("parabool", .x) ~ 2,
        TRUE ~ 4),
      .names = "{.col}_complexity"
    ),
  ) %>%
  rowwise() %>%
  mutate(
    trend = factor(
      ifelse(ndvi == bsi, "zelfde\ntrend", "verschillende\ntrend"),
      levels = c("zelfde\ntrend", "verschillende\ntrend")
    ),
    comp_bsi = factor(
      case_when(
        bsi_complexity > ndvi_complexity ~ "complexer\ndan NDVI",
        bsi_complexity < ndvi_complexity ~ "minder complex\ndan NDVI",
        TRUE ~ "zelfde complexiteit"
      ),
      levels = c("minder complex\ndan NDVI", "zelfde complexiteit",
                 "complexer\ndan NDVI")
    )
  )

# Create pairwise heat map
pairwise_trends %>%
  count(year, trend, comp_bsi) %>%
  group_by(year) %>%
  mutate(prop = n / sum(n)) %>%  # scale fill by proportion
  ungroup() %>%
  complete(year, trend, comp_bsi, fill = list(n = 0, prop = 0)) %>%
  mutate(
    is_nvt = (trend == "zelfde\ntrend" & comp_bsi != "zelfde complexiteit"),
    label = case_when(
      is_nvt ~ "n.v.t.",
      n == 0 ~ "",
      TRUE ~ paste0(n, " (", sprintf("%.3f", prop), ")")
    ),
    # Simple heuristic for readable text colour
    text_col = ifelse(prop > 0.4, "white", "black")
  ) %>%
  ggplot(aes(x = trend, y = comp_bsi)) +
  # draw normal tiles
  geom_tile(aes(fill = ifelse(is_nvt, NA, prop)), color = "white") +
  # overlay white rectangles for n.v.t. tiles
  geom_tile(data = ~filter(.x, is_nvt), fill = "white", color = "white") +
  geom_text(aes(label = label, colour = text_col), size = 3) +
  scale_colour_identity() +  # use text_col directly
  scale_fill_viridis_c(
    option = "C",
    direction = -1,
    labels = scales::percent_format(accuracy = 1),
    na.value = "white"  # ensure n.v.t. tiles are white
  ) +
  facet_wrap(~year, ncol = 2) +
  labs(x = NULL, y = NULL, fill = "Proportie") +
  theme(
    legend.position = c(0.75, 0),
    legend.background = element_rect(fill = "white", color = "darkgrey"),
    legend.margin = margin(6, 6, 6, 6),
    legend.box = "horizontal",
    legend.text = element_text(size = 8)
  )
```


```{r}
#| label: fig-trend-gewasgroep
#| fig-cap: "..."
#| fig-height: 14
# NDVI plot
p_trend_ndvi <- trend_data$ndvi %>%
  group_by(year, gwsgrp_h) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  ggplot(aes(x = gwsgrp_h, fill = trend)) +
  geom_bar(position = "fill") +
  geom_text(aes(label = n, y = 1.05), size = 2.5) +
  labs(x = "", y = "Proportie percelen", fill = "Trend in bedekking (NDVI)") +
  facet_wrap(~year, ncol = 2) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1),
        legend.position = c(0.8, -0.2),
        legend.background = element_rect(fill = "white", color = "darkgrey"),
        legend.margin = margin(6, 6, 6, 6),
        legend.box = "horizontal",
        legend.text = element_text(size = 8))
# BSI plot
p_trend_bsi <- trend_data$bsi %>%
  group_by(year, gwsgrp_h) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  ggplot(aes(x = gwsgrp_h, fill = trend)) +
  geom_bar(position = "fill") +
  geom_text(aes(label = n, y = 1.05), size = 2.5) +
  labs(x = "", y = "Proportie percelen", fill = "Trend in bedekking (BSI)") +
  facet_wrap(~year, ncol = 2) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1),
        legend.position = c(0.8, -0.2),
        legend.background = element_rect(fill = "white", color = "darkgrey"),
        legend.margin = margin(6, 6, 6, 6),
        legend.box = "horizontal",
        legend.text = element_text(size = 8))

# Combine plots
cowplot::plot_grid(
  p_trend_ndvi, p_trend_bsi,
  ncol = 1,
  labels = c("A.", "B."),
  label_size = 20,
  label_fontface = "bold",
  scale = c(0.95, 0.95)
)
```
